Quantum parallel Markov chain Monte Carlo

### 2021-11-03 main.tex
parallel compute paragraph

We propose a novel quantum computing strategy for \emph{parallel MCMC} algorithms that generate multiple proposals at each step. This strategy makes parallel MCMC amenable to quantum parallelization by untangling the generalized accept-reject step in a way that combines the Gumbel-max trick with the recent simplicial sampler proposal mechanism.  Next, we embed this classical routine within a well-known extension of Grover's quantum search algorithm to select our next Markov chain state.  Letting $D$ denote both the number of proposals and the target dimension, the combined strategy reduces the number of target evaluations required from $\order{D}$ to $\order{D^{1/2}}$  and reduces the proposal cost from $\order{D^3}$ to $\order{D^{5/2}}$.  Notably, these speedups do not preclude classical parallelization within the target evaluations and the matrix-vector multiplications of the simplicial sampler proposal.

@@

Parallel MCMC techniques use multiple proposals to obtain efficiency gains over MCMC algorithms such as Metropolis-Hastings \citep{metropolis1953equation,hastings1970monte} and its progeny that use only a single proposal.  \citet{neal2003markov} develops efficient MCMC transitions for inferring the states of hidden Markov models by proposing a `pool' of candidate states and using dynamic programming to select from among them.  \citet{tjelmeland2004using} considers inference in the general setting and shows how to maintain detailed balance for an arbitrary number (say, $K$) of proposals.  Consider a probability distribution $\pi(\dd\ttheta)$ defined on $\mathbb{R}^D$ that admits a probability density $\pi(\ttheta)$ with respect to the Lebesgue measure, i.e., $\pi(\dd \ttheta)=:\pi(\ttheta)\dd\ttheta$. To generate samples from the target distribution $\pi$, we craft a kernel $P(\ttheta_0,\dd \ttheta)$ that satisfies
\begin{align}\label{eq:fixes}
\pi(A) = \int \pi(\dd \ttheta_0) P(\ttheta_0,A) \, 
\end{align} 
for all  $A \subset \mathbb{R}^D$ for which $\pi(A) > 0$.  Letting $\ttheta_0$ denote the current state of the Markov chain, \citet{tjelmeland2004using} proposes sampling from such a kernel $P(\ttheta_0,\cdot)$ by drawing $K$ proposals $\ttheta_1,\dots,\ttheta_K$ from a distribution $Q(\ttheta_0,\dd \ttheta) =: q(\ttheta_0,\ttheta)\dd \ttheta$ and selecting the next Markov chain state from among the current and proposed states with probabilities
\begin{align}
\pi_j = \frac{\pi(\ttheta_j) \prod_{k \neq j} q(\ttheta_j,\ttheta_k)}{\sum_{i=0}^K \pi(\ttheta_j) \prod_{k \neq i} q(\ttheta_i,\ttheta_k)} \, , \quad j=0,\dots,K \, .
\end{align}
\citet{tjelmeland2004using} shows that the kernel $P(\ttheta_0,\cdot)$ constructed in such a manner maintains detailed balance and hence satisfies Equation \eqref{eq:fixes}.  Others have since built on this work, developing parallel MCMC methods that generate or recycle multiple proposals \citep{frenkel2004speed,delmas2009does,calderhead2014general,yang2018parallelizable,luo2019multiple,schwedes2021rao}.  
These developments demonstrate the ability of parallel MCMC methods to alleviate inferential challenges (such as multimodality) and to deliver performance gains over single-proposal competitors as measured by reduced autocorrelation between samples.

Nonetheless, all of them require PARALLEL COMPUTE

>>>>

We propose a novel quantum computing strategy for \emph{parallel MCMC} algorithms that generate multiple proposals at each step. This strategy makes parallel MCMC amenable to quantum parallelization by untangling the generalized accept-reject step in a way that combines the Gumbel-max trick with the recent simplicial sampler proposal mechanism.  Next, we embed this classical routine within a well-known extension of Grover's quantum search algorithm to select our next Markov chain state.  Letting $K$ denote the number of proposals in a single MCMC iteration, the combined strategy reduces the number of target evaluations required from $\order{K}$ to $\order{K^{1/2}}$.  Notably, these speedups do not preclude classical parallelization of target evaluations and proposal generations.

@@

Parallel MCMC techniques use multiple proposals to obtain efficiency gains over MCMC algorithms such as Metropolis-Hastings \citep{metropolis1953equation,hastings1970monte} and its progeny that use only a single proposal.  \citet{neal2003markov} develops efficient MCMC transitions for inferring the states of hidden Markov models by proposing a `pool' of candidate states and using dynamic programming to select from among them.  \citet{tjelmeland2004using} considers inference in the general setting and shows how to maintain detailed balance for an arbitrary number (say, $K$) of proposals.  Consider a probability distribution $\pi(\dd\ttheta)$ defined on $\mathbb{R}^D$ that admits a probability density $\pi(\ttheta)$ with respect to the Lebesgue measure, i.e., $\pi(\dd \ttheta)=:\pi(\ttheta)\dd\ttheta$. To generate samples from the target distribution $\pi$, we craft a kernel $P(\ttheta_0,\dd \ttheta)$ that satisfies
\begin{align}\label{eq:fixes}
\pi(A) = \int \pi(\dd \ttheta_0) P(\ttheta_0,A) \, 
\end{align} 
for all  $A \subset \mathbb{R}^D$ for which $\pi(A) > 0$.  Letting $\ttheta_0$ denote the current state of the Markov chain, \citet{tjelmeland2004using} proposes sampling from such a kernel $P(\ttheta_0,\cdot)$ by drawing $K$ proposals $\ttheta_1,\dots,\ttheta_K$ from a distribution $Q(\ttheta_0,\dd \ttheta) =: q(\ttheta_0,\ttheta)\dd \ttheta$ and selecting the next Markov chain state from among the current and proposed states with probabilities
\begin{align}\label{eq:probs}
\pi_j = \frac{\pi(\ttheta_j) \prod_{k \neq j} q(\ttheta_j,\ttheta_k)}{\sum_{i=0}^K \pi(\ttheta_i) \prod_{k \neq i} q(\ttheta_i,\ttheta_k)} \, , \quad j=0,\dots,K \, .
\end{align}
\citet{tjelmeland2004using} shows that the kernel $P(\ttheta_0,\cdot)$ constructed in such a manner maintains detailed balance and hence satisfies Equation \eqref{eq:fixes}.  Others have since built on this work, developing parallel MCMC methods that generate or recycle multiple proposals \citep{frenkel2004speed,delmas2009does,calderhead2014general,yang2018parallelizable,luo2019multiple,schwedes2021rao,holbrook2021generating}.  
These developments demonstrate the ability of parallel MCMC methods to alleviate inferential challenges such as multimodality and to deliver performance gains over single-proposal competitors as measured by reduced autocorrelation between samples.

But the real promise and power of parallel MCMC comes from its natural parallelizability \citep{calderhead2014general}.  Contemporary hardware design emphasizes architectures that enable execution of multiple mathematical operations simultaneously. Parallel MCMC techniques stand to leverage technological developments that keep modern computation on track with Moore's Law, which predicts that processing power doubles every two years.  For example, the algorithm of \citet{tjelmeland2004using} generates $K$ conditionally independent proposals and then evaluates the probabilities of Equation \eqref{eq:probs}.  One may parallelize the proposal generation step using parallel pseudorandom number generators (PRNG) such as those advanced in \citet{salmon2011parallel}. The computational complexity of the target evaluations $\pi(\ttheta_j)$ is linear in the number of proposals. This presents a significant burden when $\pi(\cdot)$ is computationally expensive, e.g., in big data settings, but evaluation of the target density over the $K$ proposals is again a naturally parallelizable task.  Moreover, widely available machine learning software such as \textsc{TensorFlow} allows users to easily parallelize both random number generation and target evaluations on general purpose graphics processing units (GPU) \citep{lao2020tfp}. Finally, the pairwise evaluations of the proposal density $q(\cdot,\cdot)$ in Equation \eqref{eq:probs} are $\order{K^2}$, but \citet{massive} demonstrate the natural parallelizability of such pairwise operations, obtaining multiple orders-of-magnitude speedups with contemporary GPUs.
